---
title: "The Maximum Agreement Prediction via the Concordance Correlation Coefficient"
author: "Taeho Kim^[Lehigh University, tak422@lehigh.edu], George Luta^[Georgetown University, George.Luta@georgetown.edu], Matteo Bottai^[Karolinska Institutet, matteo.bottai@ki.se], Pierre Chausse^[University of Waterloo, pchausse@uwaterloo.ca], Gheorghe Doros^[Boston University,, doros@bu.edu], Edsel A. Pena^[University of South Carolina, pena@stat.sc.edu]"
date: ""
output: rmarkdown::html_vignette
abstract: "The vignette explains how to use the malp package to compute maximum agreement prection, construct confidence intervals for the prediction and illustrate the result."
vignette: >
  %\VignetteIndexEntry{The Maximum Agreement Prediction via the Concordance Correlation Coefficient}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction


# The `malp` package

## Estimation function

Let's consider a sample $\{(X_i, Y_i)\}_{i=1}^n$ of size $n$, where
$X_i$ is a $p\times 1$ column vector and $Y_i$ is a scalar. The
maximum agreement linear predictor (MALP) for a realization $x_0$ is
defined as:

\[
\begin{split}
\hat{Y}^*(x_0)  & =  \bar{Y} + \frac{1}{\hat{\gamma}} (x_0-\bar{X})'
\mathbf{S}_\mathrm{xx}^{-1}S_\mathrm{xy}\\
& = \left(\bar{Y} -\frac{1}{\hat{\gamma}}\bar{X}'S_\mathrm{xx}^{-1}S_\mathrm{xy}
\right) + x_0'\left(\frac{1}{\hat{\gamma}}S_\mathrm{xx}^{-1}S_\mathrm{xy}
\right) \\
& = \hat\alpha_0 + x_0'\hat\alpha_1\\
& = \hat{\alpha}'X_0\,,
\end{split}
\]

where $X_0=\{1,~x_0'\}'$ and 

\begin{eqnarray*}
\bar{Y}=\frac{1}{n}\sum_{i=1}^n Y_i; & \bar{X}=\frac{1}{n}\sum_{i=1}^n X_i\\
S_\mathrm{xx} = \frac{1}{n-1}\sum_{i=1}^n (X_i-\bar{X})(X_i-\bar{X})' ;&
S_\mathrm{xy} = \frac{1}{n-1}\sum_{i=1}^n (X_i-\bar{X})(Y_i-\bar{Y})\\
\hat{\gamma} = \frac{\sqrt{S_\mathrm{YX} S_\mathrm{XX}^{-1} S_\mathrm{XY}}}{
S_\mathrm{Y}}
\end{eqnarray*}

and $S_\mathrm{Y}$ is the estimated standard deviation of $Y$. The
function `malp` return an object of class "malp". its `print` method
prints the vector $\hat{\alpha}$

```{r}
library(malp)
set.seed(112233)
X1 <- rnorm(100)
X2 <- rnorm(100)
Y <- 1+X1-X2+rnorm(100)
dat <- data.frame(Y,X1,X2)
fit <- malp(Y~X1+X2, dat)
fit
```

The coefficients can be extracted from the "malp" object 

```{r}
coef(fit)
```

## The `vcov` method

There is no asymptotic theory for the coefficients at the moment, so
the standard errors can be obtained by either the Boostrap or
Jackknife method.

```{r}
vcov(fit, "Boot", B=200)
vcov(fit, "Jackknife")
```

## The `summary'  method

This is the method to print the coefficient table with standard
errors, t-ratios and p-values. As for the `vcov` method, the standard
error can be computed using a Bootstrap or Jackknife method.

```{r}
summary(fit, vcovMet="Jackknife")
```

## The `predict` method

The `predict` method for `malp` objects is very similar to the one for
`lm` objects. The main difference is that it returns both the MALP and
the least squares prediction (LSLP). By default, it only produces the
fitted values. The result is returned as a list with two elements:
MALP and LSLP. For example, we can obtain the fitted values of MALP as
follows:

```{r}
Yhat <- predict(fit)$MALP
```

For prediction using a different set of regressors, we use the
argument `newdata`. The data must be in a 'data.frame' format and must
include all regressors included in the model. For example, if we want
the prediction for $X1=\{0\}$ and $X2=\{-0.5,0.5\}$, we first define
the new data:

```{r}
newd <- data.frame(X1=0, X2=c(-0.5, 0.5))
```

Then, we add it to the `predict` method

```{r}
predict(fit, newdata=newd)$MALP
```

To obtain the standard errors of the predictions, we set the argument
`se.fit` to TRUE. In that case, the predictions become the sub-element
`fit` of the element `MALP` of `LSLP` and the standard errors are in
the sub-element `se.fit`. Three options are currently available for
the standard error of the prediction: "Asymptotic", "Boot" or
"Jackknife". Currently, the default is the estimated asymptotic
standard error, which is only valid if the variables are jointly
normal. If the method "Boot" is selected, it is possible to change the
default number of bootstrap samples, by setting the argument 'B' to a
different integer. For example:

```{r}
pr <- predict(fit, newdata=newd, se.fit=TRUE, vcovMet="Boot", B=100)
pr$MALP$fit
pr$MALP$se.fit
```

If the argument `interval` is set to "confidence", the method returns
the prediction, the lower and upper bounds of the confidence interval. The
level of the interval is set by the argument 'level'. The default is
95%. 

```{r}
pr <- predict(fit, newdata=newd, vcovMet="As", interval="confidence", level=0.90)
pr$MALP
```

## The `plot` method

Currently, the method returns only a graph of $Y$ against the
predictions. The 45% line is added to better see the level of
agreement. 

```{r, fig.show='hold', out.width='45%'}
plot(fit, main="MALP for simulated data", cex=.3)
plot(fit, includeLS=TRUE, main="MALP and LSLP for simulated data", cex=.3)
```
