---
title: "The Maximum Agreement Prediction via the Concordance Correlation Coefficient: Applications"
author: "Taeho Kim^[Lehigh University, tak422@lehigh.edu], George Luta^[Georgetown University, George.Luta@georgetown.edu], Matteo Bottai^[Karolinska Institutet, matteo.bottai@ki.se], Pierre Chausse^[University of Waterloo, pchausse@uwaterloo.ca], Gheorghe Doros^[Boston University,, doros@bu.edu], Edsel A. Pena^[University of South Carolina, pena@stat.sc.edu]"
date: ""
bibliography: malp.bib 
output:
 pdf_document:
  number_sections: true
abstract: "This vignette reproduces the applications included in @kim2023. For more details on how to use the package, see the other vignette."
vignette: >
  %\VignetteIndexEntry{The Maximum Agreement Prediction via the Concordance Correlation Coefficient: Applications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{malp}
  %\VignetteDepends{kableExtra}
  %\VignetteDepends{xtable}
  %\VignettePackage{malp}
  %\VignetteEncoding{UTF-8}
header-includes:
- \newcommand{\trp}{^{\tt T}}
- \DeclareMathOperator{\tr}{tr}
- \newcommand{\trace}{\mbox{\rm\bf tr}}
- \DeclareMathOperator*{\argmax}{arg\,max}
- \DeclareMathOperator*{\argmin}{arg\,min}
- \newcommand{\E}{\mathrm{E}}
- \newcommand{\Var}{\mathrm{Var}}
- \newcommand{\Cov}{\mathrm{Cov}}
- \newcommand{\Cor}{\mathrm{Cor}}
- \newcommand{\CCC}{\mathrm{CCC}}
- \newcommand{\PCC}{\mathrm{PCC}}
- \newcommand{\MSE}{\mathrm{MSE}}
- \newcommand{\bs}{\boldsymbol} 
---

```{r, echo=FALSE}
library(kableExtra)
```

# Eye Data Set

In opthalmology, the central subfield macular thickness (CSMT) mea-
surements can be obtained by optical coherence tomography
(OCT). @APDS:2011 focused on two types of OCT: time-domain Stratus
OCT, the most widely used model prior to 2006; and spectral-domain
Cirrus OCT, a more advanced model. As Cirrus OCT replaces Stratus OCT
as a marker, the agreement between the measurements from two methods
is of interest to researchers in the field. For this purpose,
@APDS:2011 provided a comparison between the two approaches and
obtained a CCC-based conversion function from the Cirrus OCT to the
Stratus OCT.

In the data set, both OCTs were measured from 46 subjects, i.e., 92
eyes, but only 61\% of these observations were selected based on the
reliability of the OCTs (signal strength $\geq$ 6 for both
approaches). This subset of the original dataset is included in te
package:

```{r}
library(malp)
data(eye)
```

We can reproduce Figure 12 by first transforming the data:

```{r}
Cirrus2 <- eye$Cirrus - 60
Cirrus3 <- 0.76*eye$Cirrus - 0.51
```

We can verify that we obtain the same concordance correlation
coefficients as in the paper:

```{r}
rho <- cor(eye$Stratus, eye$Cirrus)
rhoc <- c(ccc(eye$Stratus, eye$Cirrus),
          ccc(eye$Stratus, Cirrus2),
          ccc(eye$Stratus, Cirrus3))
res <- cbind(PCC=rho, CCC=rhoc)
rownames(res) <- c("Raw", "Transform 1" , "Transform 2")
knitr::kable(res)
```

We can then reproduce Figure 12:

```{r, fig.show='hold', out.width='50%', fig.height=6, fig.width=6}
plot(Stratus~Cirrus, eye, xlab="Cirrus OCT", ylab="Stratus OCT",
     main="Penal A: Scatter Plot with Raw Data",
     xlim=c(125,325), ylim=c(125,325))
abline(1,1, lty=2)
plot(Stratus~Cirrus2, eye, xlab="Transformed Cirrus OCT",
     ylab="Stratus OCT", pch=22, xlim=c(125,325), ylim=c(125,325),
     main="Penal A: Scatter Plot with Transformed Data")
points(Cirrus3, eye$Stratus,pch=24, col=2)
abline(1,1, lty=2)
legend("topleft", c("Y=X-60","Y=0.76X-0.51"), col=1:2, pch=c(22,24),
       bty='n')
```

To produce Figure 13, we use the package to obtain the fitted lines
for both MALP and LSLP:

```{r, fig.show='hold', out.width='50%', fig.height=6, fig.width=6}
eyeOD <- subset(eye, Eye=="OD")
eyeOS <- subset(eye, Eye=="OS")
fitOD <- malp(Stratus~Cirrus, eyeOD)
plot(Stratus~Cirrus, eyeOS, xlab="Cirrus", ylab="Stratus",
     main="OD: MALP vs LSLP", xlim=c(125,325), ylim=c(100,300))
abline(fitOD, lty=1, col="blue")
abline(fitOD$lm, lty=2, col="red")
legend("topleft", c("MALP","LSLP"), col=c("blue","red"), lty=1:2,bty='n')
fitOS <- malp(Stratus~Cirrus, eyeOS)
plot(Stratus~Cirrus, eyeOS, xlab="Cirrus", ylab="Stratus",
     main="OS: MALP vs LSLP", xlim=c(125,325), ylim=c(100,300))
abline(fitOS, lty=1, col="blue")
abline(fitOS$lm, lty=2, col="red")
legend("topleft", c("MALP","LSLP"), col=c("blue","red"), lty=1:2,bty='n')
```

We reproduce here the method to generate Table 3. The first function
is to compute alll summary statistics needed (PCC, CCC and MSE):

```{r}
## Realistic
getStat <- function(obj, newD)
{
    pr <- predict(obj, newdata=newD,includeLS=TRUE)
    malp <- c(cor(newD$Stratus, pr$MALP),
              ccc(newD$Stratus, pr$MALP),
              mean((newD$Stratus-pr$MALP)^2))
    lslp <- c(cor(newD$Stratus, pr$LSLP),
              ccc(newD$Stratus, pr$LSLP),
              mean((newD$Stratus-pr$LSLP)^2))
    c(malp, lslp)    
}
```

The next function resample the data with replacement, splits the new
sample in two, use the first half to estimate the model and the second
half to compute the different statistics.

```{r}
geti <- function(dat)
{
    n <- nrow(dat)
    dat <- dat[sample(n, replace=TRUE),]
    dat1 <- dat[1:floor(n/2),]
    dat2 <- dat[-(1:floor(n/2)),]    
    fit <- malp(Stratus~Cirrus, dat1)
    getStat(fit, dat2)
}
```

The following is the bootstrap simulation for a given seed. It only
runs if the simulation file "eyeSim.rda" is not on the current
directory. 

```{r}
if (any(list.files("./") == "eyeSim.rda"))
{
    load("eyeSim.rda")
} else {
    set.seed(12052023)
    OS <- sapply(1:2000, function(i) geti(eyeOS))
    OD <- sapply(1:2000, function(i) geti(eyeOD))
    OS <- matrix(rowMeans(OS), ncol=2)
    OD <- matrix(rowMeans(OD), ncol=2)
    save(OD,OS,file="eyeSim.rda")
}
```

Finally, the following computes the results for what we refer to the
"Illustrative" case, and combine all results in one table.

```{r}
OSi <- matrix(getStat(fitOS, eyeOS), ncol=2)
ODi <- matrix(getStat(fitOD, eyeOD), ncol=2)
res <- cbind(OSi, ODi, OS, OD)
colnames(res) <- rep(c("MALP", "LSLP"),4)
rownames(res) <- c("PCC","CCC","MSE")
```

```{r, echo=FALSE, results='asis'}
t1 <- kbl(res, format="latex", digits=3)
t2 <- add_header_above(t1, c(" ", "OS" = 2, "OD" = 2, "OS" = 2, "OD" = 2))
add_header_above(t2, c(" ", "Illustrative Case" = 4, "Realistic Case" = 4)) 
```

For Figure 14, we first estimate the model with all observations. We
see that the coefficients are as shown in the paper.

```{r}
fit <- malp(Stratus~Cirrus, eye)
coef(fit)
```

To plot the confidence interval, we use the `predict` function and a
grid of new Cirrus measures. In Figure 14, all standard errors are
based on the asymptotic properties under normality.  `matplot` is used
to directly plot all lines from the `predict` output. For the
prediction intervals, we only plot the lower and upper bounds, which
correspond to the second and third row.

```{r, out.width='70%', fig.align='center'}
cir <- seq(120, 380, length.out=40)
pr <- predict(fit, newdata=data.frame(Cirrus=cir), interval="confidence",
              includeLS=TRUE)
pr2 <- predict(fit, newdata=data.frame(Cirrus=cir), interval="prediction",
              includeLS=TRUE)
plot(Stratus~Cirrus, eye, ylim=c(130,260),
     main="MALP with a 95% CI & prediction by Abedi's formula")
matplot(cir, pr$MALP, type='l',lty=c(1,2,2), col=c("blue", "gray", "gray"),
        add=TRUE)
matplot(cir, pr2$MALP[,-1], type='l',lty=3, col="green", add=TRUE)
matplot(cir, pr2$LSLP[,-1], type='l',lty=4, col="purple",add=TRUE)
abline(-.51, .76, col="red")
legend("topleft", c("MALP","Abedi","MALP CI", "MALP PI", "LSLP PI"),
       col=c("blue","red","gray","green","purple"), lty=c(1,1,2,3,4),
       bty='n', cex=.7)
```

# The BodyFat Data set

The percentage of bodyfat is an important bodily characteristic that
serves as a marker for the health status of an individual. Being able
to infer it from easily-measured or determined bodily characteristics,
such as age, weight, height, circumference measurements, or skin-fold
measurements is therefore of interest, e.g., @BW:1974, @KW:1977. For
example, one could obtain the percent bodyfat from body density via
underwater weighting based on Siri's equation $100\times
\text{Bodyfat}= 495/\text{BodyDensity}-450$ as in
@Siri:1956. Including body density (BD) and percent bodyfat
(PBF), the data set of interest contains 13 additional variables such
as age (years), weight (WGT, in pounds), height (HGT, in inches);
several body circumference measurements (in cm): neck (NCK), chest
(CST), abdomen (ABD), hip, thigh (TGH), knee (KN), ankle (ANK), biceps
(BCP), forearm (FA), and wrist (WRT) for 252 men. The data set was
originally in @Pen:1985 and became available to the public by
courtesy of Dr.\ A.\ Garth Fisher.

The dataset is included in the package and can be loaded as follows:

```{r}
data(bodyFat)
```

## Table 4

The table only returns the PCC between the different variables
included in the dataset.

```{r, message=FALSE, results='asis'}
library(xtable)
corDat <- cor(bodyFat)
print(xtable(corDat, digits=2), floating=FALSE,size='small',
      hline=c(-1,-1,0,nrow(corDat),nrow(corDat)), comment=FALSE) 
```

## Table 6

First, we estimate the five models represented by Table 5. 

```{r}
fit1 <- malp(PBF~ABD, bodyFat)
fit2 <- malp(PBF~ABD+WGT, bodyFat)
fit3 <- malp(PBF~ABD+WGT+FA+WRT, bodyFat)
fit4 <- malp(PBF~ABD+WGT+FA+WRT+Age+TGH, bodyFat)
fit5 <- malp(PBF~ABD+WGT+FA+WRT+Age+TGH+NCK+Hip, bodyFat)
```

We then organize the result in a table. The `summary` method with
`se=FALSE` is used to compute the CCC, PCC and MSE.

```{r, results='asis', results='asis'}
res <- lapply(1:5, function(i) {
    f <- get(paste("fit",i,sep=""))
    ans <- matrix(NA, 9+3, 2)
    s <- summary(f, se=FALSE)
    ans[1:length(coef(f)),] <- cbind(coef(f), coef(f$lm))
    ans[10:12,] <- cbind(s$fitMALP, s$fitLSLP)
    ans
    })
res <- do.call("cbind", res)
rownames(res) <- c(names(coef(fit5)), names(summary(fit5,se=FALSE)$fitMALP))
```

```{r, echo=FALSE, results='asis'}
options(knitr.kable.NA = '')
t1 <- kbl(res, format="latex", digits=3)
t2 <- add_header_above(t1, c(" ", rep(c("MALP","LSLP"), 5)))
add_header_above(t2, c(" ", "A"=2, "B"=2, "C"=2, "D"=2, "E"=2))
```

## Figure 15

We can use the info from the above table to produce the graph. But,
instead of having the MSE and CCC on the same graph, which requires
two different y-axes, we plot them separately.

```{r, out.width='50%', fig.show='hold'}
id <- seq(1,9,by=2)
plot(1:5, res["PCC",id], xaxt='n', main="Prediction Performance",
     xlab="Variable Subsets", ylab="CCC", ylim=c(.78,.88),
     type='l', lwd=3, col=gray(.2,.5))
axis(1, at=1:5, labels=LETTERS[1:5])
lines(1:5, res["CCC",id], col="blue", lty=2)
lines(1:5, res["CCC",id+1], col="red", lty=3)
legend("topleft", c("PCC","CCC:MALP","CCC:LSLP"),
       lty=c(1:3), col=c(gray(.2,.5),"blue","red"), cex=.75,
       bty='n')
plot(1:5, res["MSE",id], xaxt='n', main="Prediction Performance",
     xlab="Variable Subsets", ylab="MSE", ylim=range(res["MSE",]),
     type='l', lty=4, col="blue")
axis(1, at=1:5, labels=LETTERS[1:5])
lines(1:5, res["MSE",id+1], col="red", lty=5)
legend("topright", c("MSE:MALP","MSE:LSLP"),
       lty=c(4:5), col=c("blue","red"), cex=.75, bty='n')
```


## Figure 16

We can use the `plot` method form the package to produce all
scatterplots.

```{r, out.width='50%', fig.show='hold'}
plot(fit1, which="MALP", main="MALP: Subset A")
plot(fit1, which="LSLP", main="LSLP: Subset A")
plot(fit2, which="MALP", main="MALP: Subset B")
plot(fit2, which="LSLP", main="LSLP: Subset B")
plot(fit3, which="MALP", main="MALP: Subset C")
plot(fit3, which="LSLP", main="LSLP: Subset C")
plot(fit4, which="MALP", main="MALP: Subset D")
plot(fit4, which="LSLP", main="LSLP: Subset D")
plot(fit5, which="MALP", main="MALP: Subset E")
plot(fit5, which="LSLP", main="LSLP: Subset E")
```

# References




